Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-07-27 Manuscript/proto-manuscript.tex
MS: updated sections Results and discussion and Conclusions

\subsection{Electronic continuum correction for lipid bilayers}
According to the electronic continuum correction (ECC)\cite{leontyev11}, electronic
polarizability can be included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$. 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}. % that exists in living cells. 
Such a continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of ion force fields \cite{kohagen14,kohagen16,??}.

@@

The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer and an aqueous salt solutions of varying concentrations.
Water molecules were described by OPC3 model \cite{Izadi16}.
In order to test transferability of the newly developed ECC-Lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions. \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014} % cite order: [Ca_2s, Na_s, Cl_2s]
Classical molecular dynamics simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
are summarized in Table \ref{tbl:mdpar}, 
and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}. 
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

\begin{table}[btp]
  \caption{Simulation parameters}
  \label{tbl:mdpar}
  \begin{tabular}{ll}
    simulation property & parameter   \\
    \hline
    time-step           & 2~fs         \\
    equilibration time  & 100~ns  \\
    simulation time     & 200~ns  \\
    temperature         & 313~K       \\
    thermostat          & v-rescale  \cite{bussi07}   \\
    barostat            & Parrinello-Rahman, semi-isotropic \cite{parrinello81} \\
    long-range electrostatics & PME  \cite{darden93}  \\
    cut-off scheme      & Verlet \cite{Pall13}      \\
    Coulomb and VdW cut-off & 1.0~nm \\
    constraints         & LINCS, only Hydrogen atoms \cite{hess97} \\
    constraints for water & SETTLE  \cite{miyamoto92} \\
    \hline
  \end{tabular}
\end{table}

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
The code AmberTools \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.

@@

\subsection{Lipid bilayers without ions}
The scattering form factors, NMR order parameters and area per lipids calculated from 
the ECC corrected lipid model for POPC are compared to experiments and original
Lipid14 results in Fig. \ref{simVSexpNOions} and in Table~\ref{tab:apls}. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.
The structural quality is comparable to the state of art lipid models available in literature \cite{ollila16},
thus we conclude that the ECC corrected lipid model reproduces the lipid bilayer structure
in liquid disordered phase with similar accuracy than other available models
\todo{Discussion to be finished when we have all the results in the figure}.
\begin{figure}[]
  \centering
  \includegraphics[height=8.5cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
%  \caption{\label{fig:form-f}
%    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models.  }
%\end{figure}
%
%\begin{figure}[]
%  \centering
  \includegraphics[width=9.0cm]{../Fig/ipython_nb/Headgr_OPs_exp-L14-ECCL17.png}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models. 
    Headgroup and glycerol backbone order parameters from simulations with Lipid14 \cite{dickson14} and EECLipid17 models
    compared with experimental order parameters from \cite{ferreira13}.}
  \todo{Add acyl chain order parameters, POPC chemical structure}
  \todo{Should we add the results from original lipid14?}
\end{figure}
%Area per molecules extracted from MD simulations and SPD model fitted to scattering data \todo{check that this the case for the used values}
%are shown in Table \ref{tab:apls}.
\todo{finalize figure (NMR headgr. OPs + SAXS, continue in the discussion.}
\begin{table}
  \caption{Area per lipid from different models for POPC without ions\label{tab:apls} }
  \begin{tabular}{c c c}
    model          & A (Å$^2$)   & Temperature [K] \\
    \hline
    Lipid14 (literature)  & 65.6$\pm$ 0.5  &  303 \\
    Lipid14ecc0.80+sigma0.875 &        &  313    \\
    GMX small patch           & 64.9   &         \\
    GMX 4xbig patch           & 65.5   &         \\
    oMM small patch           & 63.65  &         \\
    oMM 4xbig patch           & 63.7   &         \\
    \hline
    experiment \cite{jambeck12}\todoii{REF}{put original references, not Slipids param. paper.}  & 62.7  &  293    \\
    experiment  & 64.3  &  303    \\
    experiment  & 67.3  &  323    \\
    experiment  & 68.1  &  333    \\
    experiment POPE  & 56.6 &  303    \\
    \hline
  \end{tabular}
\end{table}

Headgroup order parameter response to bound charge was evaluated against
experimental data measured with cationic surfactant
(dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) \cite{scherer89}.
The exact amount of bound charge in the membrane is known these systems, because
practically all cationic surfactant molecules are embedded in lipid bilayer due to
their amphiphilic nature. Thus, the headgroup order parameter changes as a function of
mole fraction of cationic surfactants in  Fig. \ref{OrderParameterCHANGESnewMODELS} gives
also the order parameter changes as a function of bound cations.
\todo{SAMULI: I think that we should make a separate figures for this one and ion concentrations.
  In the current figure it would be difficult to show this with a reasonable scale.
  Currently Lipid14 results are cut out, thus the scale should be increased.
  This would, however, make the changes with CaCl too small.
  The change of headgroup tilt could be incorporated in the same figure with order parameter
  response to cationic surfactant.
}
The headgroup order parameter response to bound cation concentration
is approximately linear up to $\sim$0.3 mole fraction in experiments \cite{scherer89}.
The linearity is also observed in simulations with original Lipid14 and with ECC correction.
%from experiments \cite{seelig87} and simulations with Lipid14 model and with ECC-lipids model 
%as a function of bound charge are shown in
Quantitative comparison, however, reveals that the response is overestimated in 
original Lipid14 for both segments, while the ECC corrected model gives a
good agreement for the change of $\alpha$ segment order parameter, but slightly
underestimates the $\beta$ segment order parameter change.
The overestimation of order parameter changes with original Lipid14 model 
suggests that the overestimated response with CaCl$_2$ concentration in \cite{catte16} 
can be partly explained by the high sensitivity of the headgroup order parameter response
to bound charge.

Secondly, the $S^\alpha/S^\beta$ ratio of the response is found to be slightly larger than in 
experiments (VALexp vs VALsim
\todo{Add values of $S^\alpha/S^\beta$ response for sim and experiment
  SAMULI: Maybe these could be put in the same figure with the order parameter
  response and P-N vector angle change with surfactants.
}
).
This property is kept also by the newly derived model, ECC-lipid, and, hence, 
it shows a perfect agreement in the response of the headgroup order parameter $S^\alpha$ 
but underestimates slightly the response of $S^\beta$.

\subsection{Cation binding in POPC bilayer}

Headgroup order parameter response to increasing CaCl$_2$ concentration
from experiments, original Lipid14 model and ECC corrected model are shown
in Fig. \ref{OrderParameterCHANGESnewMODELS}. The order parameter response
is significantly overestimated in original Lipid14 model, while results from
ECC corrected model are in good agreement with experiments.
This is a significant improvement over previously available models,
which always overstimate the order parameter response to CaCl$_2$
concentration \cite{catte16}. The good agreement with experiments indicate
that the binding details of Ca2+ are realistic in the ECC corrected model
and it can be thus used to study lipid-ion interaction details.

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2_surf.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Headgroup order parameter changes as a function of NaCl, CaCl$_2$ concentration and
    cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$)
    from simulations and experiments (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, surfactant \cite{scherer89}).
    Simulations with Lipid14 and Åqvist ion model from \cite{catte16,lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC1000mMCaClfiles}.
  }
  \todo{Add Lipid14-Aquist data.
    Lipid14/Åqvist data to be added from https://github.com/NMRLipids/lipid\_ionINTERACTION/blob/master/Data/POPC/CaCl/LIPID14/LIPID14caclCONSchange.dat.
    Joe: I'm not sure what is meant -- I think Aquist data are actually plotted there. I recently changed it to L14+Dang ions (both data and plot).
    I think just one ion model is sufficient (and ECC-ions are based on Dang model).
    Samuli: I think that we could add also the Aqvist data. This is in the NMRlipids II publication so it might
    make easier to follow for people who have red it publication.
  }
\end{figure*}

Ion density profiles between different simulation models are compared in Fig. \ref{fig:cacl-dens}.
Density profiles from simulations with original Lipid14 and Dang ions \cite{smith94,chang1999,dang2006} 
show a pronounced peak in the position of the phosphate moieties of POPC. 
The use of a ECC-ion model~\cite{kohagen14,kohagen16} along with original Lipid14 does not significantly change it
\todo{SAMULI: If we show density profile for this, we should show also the order parameter changes}.
The new ECC-lipid model with scaled ions exhibits on the other hand smaller density in this region 
suggesting overall weaker binding of cations (Fig.~\ref{OrderParameterCHANGESnewMODELS}). 
This demonstrates that cation binding in zwitterionic phospholipid bilayer can be accurately
described with classical MD simulation model with effective included of electronic polarizability.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities.eps}
  \caption{\label{fig:cacl-dens}
    Density profiles of \ce{Ca^{2+}} and \ce{Cl^-} for Lipid14 model with Aquist parameters
    and with ECC ions and ECCLipids17 with ECC ions. }
  \todo{SAMULI: We should add the location of bilayer here somehow.
    In NMRlipids II the location of phosphate was shown with green vertical line.}
  \todo{SAMULI: I would add Aqvist data from NMRlipids II in here as well.}
\end{figure}

Good agreement of ECC corrected model with experiments encourages us to analyse the
binding details from MD simulations. Direct analysis of contacts between ions and
lipids from simulations suggest that most common complex are ones with stoichiometry
of 2~POPC:1~\ce{Ca^{2+}}.
\todo{SAMULI: Details of this analysis have to be added.}
As shown in Fig. \ref{fig:cacl-bind} this is in agreement
with the ternary complex model suggested based on headgroup order parameter
experiments~\cite{altenbach84}.
\todo{SAMULI: The same authors have also literature, where they say that ternary complex may not
  be the only option. I will recheck and come back to this.}
%This cannot be observed in experiments without atomistic detail as it makes only a small perturabation to 
%the binding isotherm that assumes 2~POPC per 1~\ce{Ca^{2+}}. 
\todo{SAMULI: I would also analyze how much there is contact between ions and different
parts of the lipid (phosphase, carbonyl, etc.).}
In addition to the ternary complexes, there also is a non-negligable probability
of one \ce{Ca^{2+}} cross-bridging three POPC molecules
\todo{SAMULI: I think we should quantify this, i.e. how much there are these.
  Maybe also the other possible complexes? Maybe also the correlation betweem complexes
  and binding cites, if it is not too much work.}.

\todo{Finalize stoichiometry analysis for \ce{Na^+}, \ce{Ca^{2+}},
  their interaction energies with the lipid membrane, etc,
  and finalize the discussion after these results.}

@@

\begin{figure}[]
  \centering
  \includegraphics[height=9.0cm,angle=-90]{../Fig/bound-CAs_conc-eccl17.eps}
  \caption{\label{fig:cacl-bind}
    Binding isotherm assuming stoichiometry of 2~POPC:1~\ce{Ca^{2+}} as used in \cite{altenbach84} fits the simulation data nicely.}
\end{figure}

@@

%Role of water model: we use OPC3 (current best), it would be worth giving an estimate how results 
%change when we use say SPCE or even TIP3p at least in the SI (so that the reader knows what errors 
%to expect comming from these sub-optimal models). 
%In addition, there is protein force field Amber15-FB, which uses water close to OPC3, TIP3pFB \todoii{REFs}{add references here}.
%On the other hand, it might be the best if we used a model that doesn't have the dielectric constant 
%from nuclei 78, but rather 44 -- TIP4p2005 is the closest. \cite{OPC_paper, ForceBalance_paper}

%questionable charge distribution from RESP -- a leeway for further tweaks of the FF.
%It is not obvious that RESP charges provide the best description, especially due to its non-uniqe solution. 
%Shall we solve RESP fitting with the constraint of full charges and then scale down, or shall we rather 
%solve the fitting with a scaled total charge target?

\section{Conclusions}
We show that the Na+ and Ca2+ binding in phospholipid bilayers can
be accurately described with MD simulation models, where electronic
polarization is effectively included by using ECC correction.
This is a significant advantage to the other available lipid models,
which all overestimate specific cation binding affinities.  
The proposed model reproduces the lipid bilayer structural details
with similar accuracy as the other state of the art lipid models.
The correction is applied here on Lipid14 POPC model \cite{dickson14},
but we expect that the correction can be generalized also for other lipids
and force fields.
%The models were optimized to represent correct membrane structure, 
%and they were validated with the use of electrometer concept \cite{seelig87,Altenbach85,Altenbach84}.

Direct analysis of calcium binding details from MD simulations is in agreement
with ternary complex model, which is suggested based on NMR data \cite{altenbach84}.
In this model 1~calcium binds to 2~POPC molecules, which together form a ternary
complex.
%This tells us on the subtle effects of calcium in phospholipid membranes 
%possibly leading to a better understanding of their physical properties (elasticity etc.). 
%\todo{Improve this concluding-discussion with actual insights.}

This will be a foundation stone of a new open-collaboration project NPRlipid 6 in nmrlipids.blogspot.fi....

% Tables may be be put in the text as floats.
% Here is an example of the general form of a table:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
% \begin{tabular}{} command.
%
% \begin{table}
% \caption{\label{} }
% \begin{tabular}{}
% \end{tabular}
% \end{table}

>>>>

\subsection{Electronic continuum correction for lipid bilayers}
According to the electronic continuum correction (ECC)\cite{leontyev11}, electronic
polarizability can be included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}. % that exists in living cells. 
Such a continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of ion force fields \cite{kohagen14,kohagen16,??}.

@@

The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer and an aqueous salt solutions of varying concentrations.
Water molecules were described by OPC3 model \cite{Izadi16}.
In order to test transferability of the newly developed ECC-lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion model by Dang \cite{smith94,chang1999,dang2006}. 
The code AmberTools \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.

@@

\subsection{Lipid bilayers without ions}
The scattering form factors, NMR order parameters and area per lipid calculated  
using the ECC-lipids model for POPC are compared to experiments and original
Lipid14 results in Fig. \ref{simVSexpNOions} and in Table~\ref{tab:apls}. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.
The structural quality is comparable to the state of art lipid models available in literature \cite{ollila16},
thus we conclude that the ECC-lipids model reproduces the lipid bilayer structure
in liquid disordered phase with similar accuracy to other available models.

\begin{figure}[]
  \centering
  \includegraphics[height=8.5cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \includegraphics[width=9.0cm]{../Fig/ipython_nb/Headgr_OPs_exp-L14-ECCL17.png}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    Headgroup and glycerol backbone order parameters from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.}
  \todo{Add acyl chain order parameters, POPC chemical structure}
\end{figure}

\begin{table}
  \caption{Area per lipid (APL) from different models of POPC without ions\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids &        &  \\
    %Lipid14ecc0.80+sigma0.875 &        &  313    \\
    ($4.6\cdot 5.1 \, \mathrm{nm}^2$), 72 lipids patch, OPC3           & 63.2   &   313      \\
    \hline
    (6.4 nm)$^2$, 128 lipids patch, OPC3           & 64.2   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, SPC/E          & 65.1   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, OPC            & 64.4   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, TIP4p/2005     & 66.8   &  313       \\
    %($9.2\cdot 10.2 \, \mathrm{nm}^2$), 288 lipids patch           & 65.5   &  313       \\ %% not done for this model with f_sigma=0.89
    %oMM small patch           & 63.65  &         \\
    %oMM 4xbig patch           & 63.7   &         \\
    \hline
    %experiment   & 62.7  &  293    \\
    experiment \cite{jambeck12}\todoii{REF}{put original references, not Slipids param. paper.} & 64.3  &  303    \\
    experiment  & 67.3  &  323    \\
    %experiment  & 68.1  &  333    \\
    %experiment POPE  & 56.6 &  303    \\
    \hline
  \end{tabular}
\end{table}

Headgroup order parameter response to the bound charge was evaluated against
experimental data measured with the cationic surfactant
(dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) \cite{scherer89}.
The exact amount of bound charge in the membrane is known in these systems, because
practically all cationic surfactant molecules are embedded in the lipid bilayer due to
their amphiphilic nature. Thus, the changes of headgroup order parameter as a function of
mole fraction of cationic surfactants (Fig. \ref{OrderParameterCHANGESnewMODELS}) gives
also the order parameter changes as a function of the amount of bound cations.
\todo{Split the OP-changes figure in two: cationic surfactant -and- binding of cations.
      The separate figure with the cationic surfactant shall be additionally accompanied
      with a plot of Head group tilt (PN-vector?) change/response. }

The headgroup order parameter response to the bound cation concentration
is in experiments approximately linear up to $\sim$0.3 mole fraction of the cation \cite{scherer89}.
The linearity is also observed in simulations with original Lipid14 and with newly derived ECC-lipids models.
Quantitative comparison, however, reveals that the response is overestimated in 
Lipid14 for both segments, while the ECC-lipids model gives a
good agreement for the change of $\alpha$ segment order parameter, and slightly
underestimates the $\beta$ segment order parameter change.
Such a finding is not surprising, as the $\Delta S^\alpha/\Delta S^\beta$ ratio 
of the response of Lipid14 model is found to be slightly larger than in 
experiments (VALexp vs VALsim
\todo{Add values of $S^\alpha/S^\beta$ response for sim and experiment
  SAMULI: Maybe these could be put in the same figure with the order parameter
  response and P-N vector angle change with surfactants.
}
), and this property is also kept by the newly derived Lipid14-based model, ECC-lipids. 
The present comparison of headgroup order parameter changes 
suggests that the observed overestimated response of all models to 
increasing CaCl$_2$ concentration in \cite{catte16} 
can be explained at least in part by a high sensitivity of the headgroup order parameter response
to the bound charge.

\subsection{Cation binding to POPC membrane}

Headgroup order parameter response to increasing CaCl$_2$ concentration
from experiments, original Lipid14 model and ECC-lipids model are shown
in Fig. \ref{OrderParameterCHANGESnewMODELS}. The order parameter response
is significantly overestimated in original Lipid14 model, while results from
ECC-lipids model are in a good agreement with the experiments.
This is a significant improvement over current models,
which in general overstimate the head group order parameter response to CaCl$_2$
concentration \cite{catte16}. The good agreement with experiments indicate
that the binding details of \ce{Ca^2+} are realistic in the ECC-lipids model
and, hence, it can be used to study lipid-ion interaction in detail.

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2_surf.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Headgroup order parameter changes as a function of NaCl, CaCl$_2$ concentration and
    cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) mole fraction
    from simulations and experiments (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, surfactant \cite{scherer89}).
  }
\end{figure*}

Ion density profiles from simulations with different models are compared in Fig.~\ref{fig:cacl-dens}.
The density profiles from simulations with original Lipid14 \cite{dickson14} and Dang ions \cite{smith94,chang1999,dang2006} 
show a pronounced peak in the position of the phosphate moieties of POPC. 
The use of a ECC-ion model~\cite{kohagen14,kohagen16} 
or the standard Amber ion model by \AA{quist} 
along with Lipid14 does not significantly change it 
(Headgroup order parameter responses for these models in SI).
\todo{Add OP-response of Lipid14+ECC-ions into the OP-changes plot in SI}
The new ECC-lipids model with ECC-ions exhibits on the other hand smaller density in this region 
suggesting overall weaker binding of cations (Fig.~\ref{OrderParameterCHANGESnewMODELS}). 
Together with the accurate head group order parameter response (Fig.~\ref{OrderParameterCHANGESnewMODELS}),
the decrease in cation density of ECC-lipids demonstrates that 
ion binding to zwitterionic phospholipid membranes can be accurately
described with classical MD simulation models that account for electronic polarizability.

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities.eps}
  \caption{\label{fig:cacl-dens}
    Density profiles of \ce{Ca^{2+}} and \ce{Cl^-} for Lipid14 model with Dang ions
    and with ECC-ions and ECC-lipids model with ECC-ions. }
  \todo{Change the figure so that it contains a membrane background}
\end{figure}

Good agreement of ECC-lipids model with experiments encourages us to analyse the
binding details from MD simulations. Direct analysis of contacts between ions and
lipids from simulations suggest that the most abundant POPC:\ce{Ca^2+} complex 
have stoichiometry of 2~POPC:1~\ce{Ca^{2+}}.
As shown in Fig. \ref{fig:cacl-bind} this is in agreement
with the ternary complex model suggested based on headgroup order parameter
experiments~\cite{altenbach84}.
In addition to the ternary complexes, there also is a non-negligable probability
of one \ce{Ca^{2+}} cross-bridging three POPC molecules.
Technical details of the analysis are in the SI.

\todo{Put details of the cation-binding stoichiometry analysis to SI.}
\todo{Update the binding isotherm figure with new simulations}
\todo{SAMULI: The same authors have also literature, where they say that ternary complex may not
  be the only option. I will recheck and come back to this.}
\todo{SAMULI: I would also analyze how much there is contact between ions and different
parts of the lipid (phosphase, carbonyl, etc.).}
\todo{SAMULI: I think we should quantify this, i.e. how much there are these.
  Maybe also the other possible complexes? Maybe also the correlation betweem complexes
  and binding sites, if it is not too much work.
      JOE: This looks like a careful work for the next paper to me. 
           I'd only add a relatively simple analysis of binding sites and probably the propensity of 1-2-3 membered clusters.}. 
\todo{Finalize stoichiometry analysis for \ce{Na^+}, \ce{Ca^{2+}},
  their interaction energies with the lipid membrane, etc,
  and finalize the discussion after these results.}

@@

\begin{figure}[]
  \centering
  \includegraphics[height=9.0cm,angle=-90]{../Fig/bound-CAs_conc-eccl17.eps}
  \caption{\label{fig:cacl-bind}
    Binding isotherm assuming stoichiometry of 2~POPC:1~\ce{Ca^{2+}} as used in \cite{altenbach84} fits our simulation data well .}
\end{figure}

@@

\section{Conclusions}
We show that the \ce{Na^+} and \ce{Ca^2+} binding in phospholipid bilayers can
be accurately described with classical MD simulation models, where electronic
polarization is effectively included by using electronic continuum correction.
This is a significant improvement over other available lipid models,
which all overestimate specific cation binding affinities \cite{catte16}.  
The newly proposed model, which we denote as ''ECC-lipids~17'', 
reproduces the lipid bilayer structural details
with similar accuracy as other state of the art lipid models.
ECC-lipids~17 exhibit accurate headgroup order parameter response to
bound cations, monovalent \ce{Na^+} and cationic surfactant dihexadecyldimethylammonium bromide, 
and divalent \ce{Ca^2+}.
Several water models 
(OPC3\cite{Izadi16}, OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}) 
were used to exemplify the transferability of 
the parameters of the new ECC-lipids~17 force field.

Direct analysis of calcium binding details from MD simulations is in agreement
with ternary complex model, which is suggested based on NMR data \cite{altenbach84}.
In this model 1~calcium binds to 2~POPC molecules, which together form a ternary
complex.

The electronic continuum correction is applied here on Lipid14 POPC model \cite{dickson14},
but we expect that the correction can be generalized also for other lipids
and force fields.

This work will serve as a foundation stone of a 
new open-collaboration project NMRlipids~VI in nmrlipids.blogspot.fi.
